### 2022-03-05 iwsm/abstract.tex
4 pages

\abstract{
Labelled networks are an important class of data,
naturally appearing
in numerous applications in science and engineering.
A typical inference goal is to determine how the vertex labels
(or {\em features}) affect the network's structure.
In this work, we introduce a new generative model, the feature-first block model (FFBM),
that facilitates the use of rich queries on labelled networks.
We develop a Bayesian framework and devise a two-level Markov chain Monte 
Carlo approach to efficiently sample from the
relevant posterior distribution of the FFBM parameters. This allows us to infer if and how the observed vertex-features affect macro-structure.
We apply the proposed methods to a variety of network data
to extract the most important features along which the vertices
are partitioned. The main advantages of the proposed approach are
that the whole feature-space is used automatically and that features
can be rank-ordered implicitly according to impact.
}

>>>>

\abstract{
Labelled networks are an important class of data,
naturally appearing
in numerous applications in science and engineering.
A typical inference goal is to determine how the vertex labels
(or {\em features}) affect the network's structure.
In this work, we introduce a new generative model, the feature-first block model (FFBM),
that facilitates the use of rich queries on labelled networks.
We develop a Bayesian framework and devise a two-level Markov chain Monte 
Carlo approach to efficiently sample from the
relevant posterior distribution of the FFBM parameters. This allows us to infer if and how the observed vertex-features affect macro-structure.
We apply the proposed methods to several real-world networks
to extract the most important features along which the vertices
are partitioned. The approach stands out from its peers in
that the whole feature-space is used automatically and features
can be rank-ordered implicitly by importance.
}

### 2022-03-05 iwsm/conclusion.tex
4 pages

The proposed Feature-First Block Model (FFBM)
is a new generative model for labelled networks.
It is a hierarchical Bayesian model, 
well-suited for describing how features
affect network structure.
The Bayesian inference tools developed in this 
work facilitate the identification of 
vertex features that are in some way 
correlated with the network's graphical structure.
Consequently, 
finding the features that best describe the most 
pronounced partition, makes it possible in practice
to examine the existence of -- and to make a case for --
causal relationships.

>>>>

%The proposed Feature-First Block Model (FFBM)
%is a new generative model for labelled networks.
%It is a hierarchical Bayesian model, 
%well-suited for describing how features
%affect network structure.
%The Bayesian inference tools developed in this 
%work facilitate the identification of 
%vertex features that are in some way 
%correlated with the network's graphical structure.
%Consequently, 
%finding the features that best describe the most 
%pronounced partition, makes it possible in practice
%to examine the existence of -- and to make a case for --
%causal relationships.

### 2022-03-05 iwsm/experiments.tex
4 pages

We apply our proposed methods to a variety of real-world datasets.
For reference, the inferred partitions for all of these are given on Figure~\ref{fig:graphs-all}.
%We employ the following metrics to assess model performance. 
%First, the average
%description length per entity (nodes and edges) 
%$\bar{S}_e$ 
%used to gauge the SBM fit is defined as:
%%
%\begin{equation}
%	\bar{S}_e \triangleq \frac{1}{(N+E) |\mathcal{T}_b|} \sum_{t\in \mathcal{T}_b} S \left( b^{(t)} \right).
%	\label{eqn:mean-dl}
%\end{equation}
%%
%Next, to assess the performance of the feature-to-block predictor, 
%the vertex set $[N]$ 
%is partitioned at random so that 
%a constant fraction $f$ of vertices form the training set $\mathcal{G}_0$ and 
%the remainder form the test set $\mathcal{G}_1$.
%The $b$-chain is run using the whole network but only vertices $v \in \mathcal{G}_0$ are 
%used for the $\theta$-chain. 
%Then the  the average cross-entropy loss 
%over each set is used to gauge the quality of the fit,
%%
%\begin{equation}
%	\bar{\mathcal{L}}_\star \triangleq \frac{1}{|\mathcal{T}_\theta|} \sum_{t \in \mathcal{T}_\theta} \mathcal{L}_\star^{(t)},
%	\quad \textrm{where} \quad
%	\mathcal{L}_\star^{(t)} \triangleq \frac{1}{|\mathcal{G}_\star|} \sum_{i \in \mathcal{G}_\star}\sum_{j \in [B]} \hat{y}_{ij} \log \frac{1}{\phi_j \left(x_i; \theta^{(t)} \right)},
%	\label{eqn:cross-entropy-loss}
%\end{equation}
%%
%where $\star \in \{0, 1\}$ toggles between the training and test sets
%and $\hat{y}_{ij}$ is defined in~(\ref{eqn:y-hat}).
%%
%Nevertheless, the cross-entropy loss is a coarse measure of fit. 
%A new measure, specific to each detected block,
%can be defined as follows. Let
%$\mathcal{B}_\star(j)$ 
%be the set of vertices with maximum a posteriori probability of belonging 
%to block~$j$,
%$
%	\mathcal{B}_\star(j) \triangleq \{i \in \mathcal{G}_\star : \hat{b}_i = j\},
%$
%where
%$ 
%	\hat{b}_i \triangleq \underset{j}{\operatorname{argmax}}\hat{y}_{ij},
%$
%and
%define the {\em block-accuracy} for block $j$ as,
%%
%\begin{equation}
%	\eta_\star(j) \triangleq \frac{1}{|\mathcal{B}_\star (j)| \cdot 
%	|\mathcal{T}_\theta| } 
%	\sum_{i \in \mathcal{B}_\star (j)}  \sum_{t \in \mathcal{T}_\theta}
%	\boldsymbol{1} \left\{\hat{b}_i = \underset{j}{\operatorname{argmax}}\phi_j \left( x_i; \theta^{(t)} \right) \right\}.
%	\label{eqn:accuracy}
%\end{equation}
%%
%This effectively tests whether the feature-to-block and 
%graph-to-block predictions agree in their largest component.
%For the higher-dimensional datasets, we also apply the 
%dimensionality reduction method 
%of Section~\ref{sec:dim-reduction}.  
%We then retrain the feature-block predictor using only the retained 
%feature set $\mathcal{D}'$, and report the log-loss over the training and 
%test sets for the reduced classifier -- 
%denoted $\bar{\mathcal{L}}_0'$ and $\bar{\mathcal{L}}_1'$ respectively.

>>>>

We apply our proposed methods to a variety of real-world datasets.
For reference, the inferred partitions $b$ for all of these are given on Figure~\ref{fig:graphs-all}.
We require metrics to assess model performance. 
First, the average
description length per entity (nodes and edges) 
$\bar{S}_e$ 
is used to gauge the SBM fit.
%%
%\begin{equation}
%	\bar{S}_e \triangleq \frac{1}{(N+E) |\mathcal{T}_b|} \sum_{t\in \mathcal{T}_b} S \left( b^{(t)} \right).
%	\label{eqn:mean-dl}
%\end{equation}
%%
Second, to assess the performance of the feature-to-block predictor, 
the vertex set $[N]$ 
is partitioned at random into training and test sets, $\mathcal{G}_0$ and $\mathcal{G}_1$. 
Then the average cross-entropy loss 
over each set is used to gauge the quality of the fit, $\bar{\mathcal{L}}_\star$,
%%
%\begin{equation}
%	\bar{\mathcal{L}}_\star \triangleq \frac{1}{|\mathcal{T}_\theta|} \sum_{t \in \mathcal{T}_\theta} \mathcal{L}_\star^{(t)},
%	\quad \textrm{where} \quad
%	\mathcal{L}_\star^{(t)} \triangleq \frac{1}{|\mathcal{G}_\star|} \sum_{i \in \mathcal{G}_\star}\sum_{j \in [B]} \hat{y}_{ij} \log \frac{1}{\phi_j \left(x_i; \theta^{(t)} \right)},
%	\label{eqn:cross-entropy-loss}
%\end{equation}
%%
%where $\star \in \{0, 1\}$ toggles between the training and test sets.
%
%and $\hat{y}_{ij}$ is defined in~(\ref{eqn:y-hat}).
%
%Nevertheless, the cross-entropy loss is a coarse measure of fit. 
%A new measure, specific to each detected block,
%can be defined as follows. Let
%$\mathcal{B}_\star(j)$ 
%be the set of vertices with maximum a posteriori probability of belonging 
%to block~$j$,
%$
%	\mathcal{B}_\star(j) \triangleq \{i \in \mathcal{G}_\star : \hat{b}_i = j\},
%$
%where
%$ 
%	\hat{b}_i \triangleq \underset{j}{\operatorname{argmax}}\hat{y}_{ij},
%$
%and
%define the {\em block-accuracy} for block $j$ as,
%%
%\begin{equation}
%	\eta_\star(j) \triangleq \frac{1}{|\mathcal{B}_\star (j)| \cdot 
%	|\mathcal{T}_\theta| } 
%	\sum_{i \in \mathcal{B}_\star (j)}  \sum_{t \in \mathcal{T}_\theta}
%	\boldsymbol{1} \left\{\hat{b}_i = \underset{j}{\operatorname{argmax}}\phi_j \left( x_i; \theta^{(t)} \right) \right\}.
%	\label{eqn:accuracy}
%\end{equation}
%%
%This effectively tests whether the feature-to-block and 
%graph-to-block predictions agree in their largest component.
%
For the higher-dimensional datasets, we also develop and apply a novel 
dimensionality reduction method to select only the top $D'$ features.  
We then retrain the feature-block predictor using only the retained 
feature set, and report the cross-entropy loss over the training and 
test sets for the reduced classifier -- 
denoted $\bar{\mathcal{L}}_\star'$.

Table~\ref{tab:results} summarises the high-level results for each experiment. 
We see that the dimensionality reduction procedure 
brings the training and test losses closer. This indicates that
the retained features
are indeed well correlated with the underlying graphical 
partition and that the approach generalises correctly. 
%The test loss variance is higher than the training loss variance as the test set is smaller and so more susceptible to variability in its construction.
%%
%The average description length per entity of the graph,
%$\bar{S}_e$,
%has very small variance, suggesting that
%the detected communities can be found reliably (to within an arbitrary 
%relabelling of blocks).

### 2022-03-05 iwsm/ffbm.tex
4 pages

In this section, we propose a novel generative model for labelled networks. We call this the feature-first block model (FFBM),
illustrated in Figure~\ref{fig:ffbm}.
%
Let $N$ denote the number of vertices, $B$ the number of blocks
and ${\cal X}$ the set of values each feature can take.
%
%We define the vector $x_i \in \mathcal{X}^D$ as the feature vector for vertex $i$, 
%where $D$ is the number of features associated with each vertex.
%For example, in the datasets we analyse, we deal with binary feature flags
%(denoting the presence/absence of each feature),
%so $\mathcal{X} = \{0, 1\}$.
%
We write $X$ for the $N\times D$ {\em feature matrix} containing
the feature vectors $\{x_i\}_{i=1}^{N}$ 
as its rows.

@@

Once the block memberships $b$ have been generated, we then draw the 
adjacency matrix of the graph, $A$, from the microcanonical DC-SBM [Peixoto (2017)] with additional parameters 
$\psi$,
%
\begin{equation}
	A \sim \textrm{DC-SBM}_{\textrm{MC}} (b, \psi).
	\label{eqn:A-generation}
\end{equation}

>>>>

We propose a novel generative model for labelled networks -- which we call the feature-first block model (FFBM),
illustrated in Figure~\ref{fig:ffbm}.
%
Let $N$ denote the number of vertices, $B$ the number of blocks
and ${\cal X}$ the set of values each feature can take.
%
%We define the vector $x_i \in \mathcal{X}^D$ as the feature vector for vertex $i$, 
%where $D$ is the number of features associated with each vertex.
%For example, in the datasets we analyse, we deal with binary feature flags
%(denoting the presence/absence of each feature),
%so $\mathcal{X} = \{0, 1\}$.
%
We write $X$ for the $N\times D$ {\em feature matrix} containing
the feature vectors $\{x_i\}_{i=1}^{N}$ 
as its rows.

@@

Once the block memberships $b$ have been generated, we then draw the 
adjacency matrix of the graph, $A$, from the microcanonical DC-SBM, Peixoto (2017), with additional parameters 
$\psi$,
$
	A \sim \textrm{DC-SBM}_{\textrm{MC}} (b, \psi).
	\label{eqn:A-generation}
$

### 2022-03-05 iwsm/inference.tex
4 pages

Having completed the definition of the FFBM, we wish to leverage it 
to perform inference. Specifically, given a labelled network $(A, X)$, we wish to infer if and how the observed features $X$ impact the graphical structure $A$. Formally,
this means characterising the posterior distribution:
$
p(\theta|A, X) \propto p(\theta) \cdot p(A | X, \theta).
$
%Although the prior is easily computable, 
%computing the likelihood $p(A| X. \theta)$
%requires summing over all $B^N$ latent block-states which is 
%clearly impractical.
%In fact, this
%approach is doubly intractable as we would also 
%need to compute the normalising constant $p(A|X)$.
Therefore, following standard Bayesian practice,
instead we aim to draw samples from the posterior,
%
\begin{equation}
	\label{eqn:theta-target}
	\theta^{(t)} \sim p(\theta | A, X).
\end{equation}
%
We propose an iterative Markov chain Monte Carlo
(MCMC) approach to obtain these samples
$\{\theta^{(t)}\}$. We first draw a sample $b^{(t)}$ 
from the block membership posterior,
and then use $b^{(t)}$ to obtain a corresponding
sample $\theta^{(t)}$:
%
\begin{equation}
	b^{(t)} \stackrel{\rm distr}{\approx} p \big( b | A, X \big) 
	\quad \textrm{then} \quad
	\theta^{(t)} \stackrel{\rm distr}{\approx} 
	p\big(\theta | X, b^{(t)} \big),
\end{equation}
%
%where these approximations become exact as
%the number of MCMC iterations $t\to\infty$.
Splitting the Markov chain into two levels side-steps the intractable summation over
all latent $b \in [B]^N$ required to directly compute the likelihood, $p(A| X, \theta)$.
The resulting $\theta^{(t)}$ samples are asymptotically
unbiased in that the expectation of 
their distribution converges to the true posterior.
%%
%\begin{equation}
%\lim_{t\to\infty}
%\mathbb{E}_{b^{(t)}} \left[p \left( \theta | X, b^{(t)} \right) \right] = \sum_{b \in [B]^N} p(\theta | X, b) p(b | A, X) = p(\theta | A, X).
%\label{eqn:theta-unbiased}
%\end{equation}
%%
%This is an example of a pseudo-marginal approach;
%see, e.g., Andrieu and Roberts (2009) 
%for a detailed rigorous derivation based on~(\ref{eqn:theta-unbiased}).
%
\begin{figure}[!ht]
	\centering

@@

Figure \ref{fig:samp-sequence} shows an overview of the proposed method, with $q$ and $\alpha$ denoting the Metropolis-Hastings proposal distribution and acceptance probability respectively.
Due to the FFBM's formulation, evaluating $p(b| X)$ does not depend on $X$ so
we do not need $X$ to sample $b$.
And on the other level, in order to obtain 
samples for $\theta$
we use only $b$ but not $A$, as $(\theta \perp \!\!\! \perp A )| b$.

>>>>

We wish to leverage the FFBM for inference. Specifically, given a labelled network $(A, X)$, we wish to infer if and how the observed features $X$ impact the graphical structure $A$. Formally,
this means characterising the posterior distribution:
$
p(\theta|A, X) \propto p(\theta) \cdot p(A | X, \theta).
$
%Although the prior is easily computable, 
%computing the likelihood $p(A| X. \theta)$
%requires summing over all $B^N$ latent block-states which is 
%clearly impractical.
%In fact, this
%approach is doubly intractable as we would also 
%need to compute the normalising constant $p(A|X)$.
Therefore, following standard Bayesian practice,
instead we aim to draw samples from the posterior,
$
	\label{eqn:theta-target}
	\theta^{(t)} \sim p(\theta | A, X).
$
We propose an iterative Markov chain Monte Carlo
(MCMC) approach to obtain these samples
$\{\theta^{(t)}\}$. We first draw a sample $b^{(t)}$ 
from the block membership posterior,
and then use $b^{(t)}$ to obtain a corresponding
sample $\theta^{(t)}$:
%
\begin{equation}
	b^{(t)} \stackrel{\rm distr}{\approx} p \big( b | A, X \big) 
	\quad \textrm{then} \quad
	\theta^{(t)} \stackrel{\rm distr}{\approx} 
	p\big(\theta | X, b^{(t)} \big),
\end{equation}
%
%where these approximations become exact as
%the number of MCMC iterations $t\to\infty$.
Splitting the Markov chain into two levels side-steps the intractable summation over
all latent $b \in [B]^N$ required to directly compute the likelihood, $p(A| X, \theta)$.
The resulting $\theta^{(t)}$ samples are asymptotically
unbiased in that the expectation of 
their distribution converges to the true posterior.
%%
%\begin{equation}
%\lim_{t\to\infty}
%\mathbb{E}_{b^{(t)}} \left[p \left( \theta | X, b^{(t)} \right) \right] = \sum_{b \in [B]^N} p(\theta | X, b) p(b | A, X) = p(\theta | A, X).
%\label{eqn:theta-unbiased}
%\end{equation}
%%
%This is an example of a pseudo-marginal approach;
%see, e.g., Andrieu and Roberts (2009) 
%for a detailed rigorous derivation based on~(\ref{eqn:theta-unbiased}).
%
\begin{figure}[!ht]
	\centering

@@

Figure \ref{fig:samp-sequence} shows an overview of the proposed method, with $q$ and $\alpha$ denoting the Metropolis-Hastings proposal distribution and acceptance probability.
Due to the formulation of the FFBM, evaluating $p(b| X)$ does not depend on $X$ so
we do not need $X$ to sample $b$.
And on the other level, in order to obtain 
samples for $\theta$
we use only $b$ but not $A$, as $(\theta \perp \!\!\! \perp A )| b$.

### 2022-03-05 iwsm/introduction.tex
4 pages

Many real-world networks exhibit strong community structure, with most nodes belonging to densely connected clusters. 
%Finding ways to recover the latent communities from the observed graph is an important
%task in many applications,
%including
%compression [Abbe (2016)] and link prediction 
%[Zhu et al. (2016)].
In this work, we examine vertex-labelled networks, 
referring to the labels as {\em features}. A typical goal is to determine whether a given feature impacts graphical structure. Answering this requires a random graph model;
the standard is the stochastic block model (SBM) [Nowicki and Snijders (2001)].
%Numerous variants of the SBM  have been proposed, e.g., the MMSBM [Airoldi et al. (2009)]
%and OSBM [Zhu et al. (2016)], but these do not include features in the graph generation
%process.

INSERT REBUTTAL OF CURRENT METHODS

To analyse a labelled network using one of the simple SBM variants, a typical procedure would be to partition the graph into blocks grouped by distinct values of the feature of interest. The associated model can then be used to test for evidence of heterogeneous connectivity between the feature-grouped blocks. Nevertheless, this approach can only consider disjoint feature sets and the feature-grouped blocks are often an unnatural partition of the graph.

We would instead prefer to partition the graph into its most natural blocks and then find which of the available features -- if any -- best predict the resulting partition. Thus motivated, we present a novel framework for modelling labelled networks, which we call the feature-first block model (FFBM). This is an extension of the SBM to labelled networks.

>>>>

Many real-world networks exhibit strong community structure, with most nodes belonging to densely connected clusters. 
%Finding ways to recover the latent communities from the observed graph is an important
%task in many applications,
%including
%compression [Abbe (2016)] and link prediction 
%[Zhu et al. (2016)].
In this work, we examine vertex-labelled networks, 
referring to the labels as {\em features}. A typical goal is to determine whether a given feature impacts graphical structure. Answering this requires a random graph model;
the standard is the stochastic block model (SBM) -- see Peixoto (2017).
%Numerous variants of the SBM  have been proposed, e.g., the MMSBM -- Airoldi et al. (2009) --
%and OSBM -- Zhu et al. (2016) -- but these do not include features in the graph generation
%process.

Analysing a labelled network with a simple SBM variant, requires partitioning the graph into blocks grouped by distinct values of the feature of interest. The associated model can then be used to test for evidence of heterogeneous connectivity between the feature-grouped blocks. Nevertheless, this approach can only consider disjoint feature sets and the feature-grouped blocks are often an unnatural partition of the graph.

We would instead prefer to partition the graph into its most natural blocks and then find which of the available features -- if any -- best predict the resulting partition. Thus motivated, we present a novel framework for modelling labelled networks.
This would not the first extension of the SBM to labelled networks, e.g. Stanley et al (2019). However, most of the current approaches are focused on leveraging feature information to partition the graph more reliably in the presence of noise.
We seek instead to develop a model well suited for inferring how vertex features impact graphical structure and to report our confidence in those conclusions.
